---
title: "lab_03"
author: "Derek Willis/ Rob Wells"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load 
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008) |>
  clean_names()

# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |>
  clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. 

Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

#A1:I found 86 records.

```{r}
payments |>
  filter(str_detect(vendor_name, "PRINCE GEORGE") & (str_detect(vendor_name, "SCHOOL") | str_detect(vendor_name, "EDUCATION") | str_detect(vendor_name, "PGCPS"))) |>
  group_by(vendor_name) |>
  summarise(count = n()) |>
  arrange(vendor_name)
```
```{r}

pg_schools <- payments |>
  filter((str_detect(vendor_name, "PRINCE GEORGE") & str_detect(vendor_name, "SCHOOL")) | (str_detect(vendor_name, "PRINCE GEORGE") & str_detect(vendor_name, "EDUCATION")) | str_detect(vendor_name, "PRINCE GEORGE") & (str_detect(vendor_name, "BOARD OF EDUCATION")) | str_detect(vendor_name, "PGCPS") | str_detect(vendor_name, "PG COUNTY SCHOOL"))

nrow(pg_schools)
```
Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

#A2: Fiscal year 2024 had the largest total amount ($1,782,519,925).  The payments show a general upward trend over time and the most recent years (2019-2024) have the highest amounts, all exceeding $1.3 billion annually. There's a notable jump from 2023 to 2024 of over $200 million. FY 2025, with its abismally low levels -- even lower than FY 2012, stands out to me. This is a direct result of multiple cuts, including but not limited to:
##1. cut in federal funds - https://marylandmatters.org/2025/07/10/federal-freeze-on-funds-could-cost-maryland-schools-125-million-this-year/
##2. cuts to the 'Blueprint for Maryland’s Future' - https://cnsmaryland.org/2025/04/01/maryland-wrestles-with-blow-to-the-state-schools-budget/
##3. Dip in general fund revenues - https://mdeconomy.org/budget-deal-leaves-education-funding-state-services-deeply-uncertain/
##4. The U.S. Department of Education freezing the spending of COVID-19 recovery funds (about $400M) - https://marylandmatters.org/2025/03/31/maryland-faces-418m-catastrophic-loss-in-pandemic-era-relief-funds/

```{r}
by_fiscal_year <- pg_schools |>
  group_by(fiscal_year) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```

Q3. This question is in two parts. 

Part 1: Code: You're interested in examining the vendors paid for capital projects at the University of Maryland. Write code that does the following: 
--Filters payments data to capital projects at the University of Maryland
--Produce a list of vendors and the total amount they received. 

Part 2: Provide a written answer, not code.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? 
--Provide at least three examples of vendor names that are similar and should be combined into one. 
--Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?


#A3. The data requires significant cleaning to combine similar vendor names. Examples of vendors that should be combined (non-exhaustive):

###'WHITING TURNER CONTRACTING CO' ($509,838,206) and 'WHITING TURNER CONTRACTING COMPANY' ($478,385,445) - same company, different abbreviations of "COMPANY"
###'A MORTON THOMAS & ASSOC INC' ($7,320.00) and 'A MORTON THOMAS & ASSOCIATES INC' ($87,132.95) - same company, "ASSOC" vs "ASSOCIATES"
###'A&I  INC' (with double space) - $65,019.25 and 'A&I INC.' (single space, with period) - $146,226.25
###'AEROSOL MONITORING & ANALYSIS INC' ($15,988.58) and 'AERSOL MONITORING & ANALYSIS INC' ($500.00) - same company, spelling error "AERSOL" vs "AEROSOL"

## To clean this data I would:
### str_to_upper() to standardize capitalization (from pre_lab_02)
### str_replace() to remove periods and spaces: str_replace(vendor_name, "\.", "")
### str_replace() to standardize abbreviations: str_replace(vendor_name, " ASSOC ", " ASSOCIATES ") 
### clean_names() could be used but I am not sure how to use it for values within a column
## There could be more errors and more functions needed.


```{r}
umd_capital_proj <- payments |>
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |>
  group_by(vendor_name) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))

head(umd_capital_proj, 20)
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. 

Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. 
Then create a total amount of money for each fiscal year. 

Write a sentence summarizing how stem cell research funding has changed over time. Be specific with the totals.

#A4:Stem cell research funding started high in 2009 at $14,643,721, then generally declined through from 2010-2020 and it saw a notable dip to $4,832,225 in 2021, followed by a recovery to $7,069,212 in 2022 and a significant jump to $12,415,121 in 2023.

```{r}
md_grants_loans |>
filter(str_detect(str_to_upper(description), "STEM CELL")) |>
  group_by(description) |>
  summarise(count = n()) |>
  arrange(description)
```
```{r}
stem_cell_funding <- md_grants_loans |>
  filter(str_detect(str_to_upper(description), "STEM CELL")) |>
  group_by(fiscal_year) |>
  summarise(total_amount = sum(amount)) |>
  arrange(fiscal_year)

stem_cell_funding
```
Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

#A5:There are 21 unique grantees in the 20742 zip code. The data requires cleaning to combine 'University of Maryland College Park' and 'UNIVERSITY OF MARYLAND'. OVALTZ INC stands out since its existence could not be established after my research.

##To clean this data I would use str_detect () function to club all University of Maryland grantees. 

###Note to self:
####Haystack Solutions, Inc. founded 2016 has 2 employees. 
####Remedium Technologies, a medical device company and University of Maryland spin-out advancing the standard of care for treating uncontrolled hemorrhage, announces the 510(k) clearance of its Hemogrip™ Patch by the U.S. Food and Drug Administration (FDA).
####Zymetis, founded in 2006 by Dr. Steve Hutcheson of University of Maryland, is a biotechnology company working to produce affordable"next-generation transportation fuels." Google Maps shows that it has permanently closed.In 2011, Aemetis acquired Zymetis, Inc.


Hemogrip™ Patch, the company's lead product, is indicated for use at vascular access sites, where it rapidly controls bleeding that occurs when accessing veins or arteries for various medical treatments and applications.

```{r}
umd_area_grants <- md_grants_loans |>
  filter(zip_code == "20742" | str_detect(zip_code, "20742")) |>
  group_by(grantee) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))

nrow(umd_area_grants)
umd_area_grants
```
