---
title: "lab_07"
author: "Derek Willis/Rob Wells"
date: "2025-10-21"
output: html_document
---


Lab07 - API Keys and Tidycensus


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

install.packages("tidycensus")

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Security and API keys

You'll need your Census API Key. Here's how to do that securely:

r_environ is a hidden file that can securely store API keys
```{r}
usethis::edit_r_environ()
# a new text file opens in RStudio
```

1. Add your API keys in the r_environ file
CENSUS_API_KEY=XXXXX
OPENAI_API_KEY=XXXX
etc....

2. Save the r_environ() file and RESTART R

# Hide your API Key from GitHub

```{r}
# Check if .gitignore exists in your project
file.exists(".gitignore")
```

Assuming you don't have a .gitignore file, then create one
1. RStudio:
- File → New File → Text File

2. ## Add These Lines to Your `.gitignore`
# R files
.Rhistory
.RData
.Rproj.user/
.Renviron
.Rprofile

# Environment files
.env
*.env

# API keys and secrets
*secret*
*password*
*api_key*

3. save the file a .gitignore in the root directory of your project
yes, you have to precede the file with a dot: .gitignore

4. Veryify it is working

```{r}
system("git status")
```


## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
census_api_key("CENSUS_API_KEY")
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2023 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2023 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
sf3 <- load_variables(2000, "sf3", cache = TRUE)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2023 5-year estimates? Hint: look at the variables for ACS in 2023 and find the variable for median age. **A1**

```{r}
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe created in Q1 to answer this, but you must write code to display only those counties with a median age of at least 50. **A2**

```{r}
```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1.  Load the Montgomery County 911 overdose calls data.
```{r}

```

2.  Create a new dataframe with the total number of calls for each zip code.
```{r}

```

3.  Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one). Then get from the ACS data from all Maryland zip codes using that variable - you will need to specify an earlier year to get the data.
```{r}

```

4.  Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
```{r}

```

5.  Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.
```{r}

```


Which zip code has the highest rate of calls? Which has the lowest?

**A3**

```{r}
```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4**

**Write your answer here:**

